---
title: "Enrichment analysis with clusterProfiler"
author: "Rene Welch"
date: February, 2018
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r start,include=FALSE,echo=FALSE,eval=TRUE}
## loads packages and data

library(org.Hs.eg.db)
library(tidyverse)
library(DESeq2)
library(clusterProfiler)
library(gridExtra)

fdr = 5e-2

knitr::opts_chunk$set(include = FALSE,echo = FALSE,eval = TRUE)

load(here::here("apps/Fig3_dashboad/fig3data.RData"))

```

# Introduction

We are going to perform a pathway analysis, I am using the data from Figure 3, i.e. genes that were differentiallly expressed in specific EBV or NOKS cell lines after applying the MC  treatment:

```{r data,include=TRUE}
ls()
```

Mark already analyzed the data using the TPM matrix and GSEA, so I am going to use `r BiocStyle::Biocpkg("clusterProfiler")` as an alternative method using the data already processed by `r BiocStyle::Biocpkg("DESeq2")`. 

As far as I can tell, the advantages of this package is that:

1. It can perform pathway analysis using the Hallmark gene data set from **MolSigDB**

2. It contains several visualization methods, including some based on GSEA.

3. This may be minor, but for ChIP-seq annotation we have used `r BiocStyle::Biocpkg("ChIPseeker")` which was created by the same author.

# Preparing the data

## Hallmark gene list V5.1

We downloaded the **MolSigDB** curated gene sets from [the GSEA website](http://software.broadinstitute.org/gsea/downloads.jsp), and then we load the hallmark gene set from **MolSigDB**, using the 5.1 version:

```{r hallmark,cache=TRUE,include=TRUE,message=FALSE,warning=FALSE}

gmt_files = here::here("data/Pathway") %>% 
  list.files(full.names = TRUE,recursive = TRUE) %>% 
  str_subset("v5.1") %>% 
  str_subset(".gmt") %>% 
  str_subset("h.all") %>% 
  set_names(
    {
      basename(.) %>% 
        str_replace(".gmt","") %>% 
        str_split("\\.") %>% 
        map_chr( ~ .[length(.)])
    }
  )


hallmark = gmt_files %>% 
  map(read.gmt)

gmt_files %>% 
  basename()

```

## Diff expressed genes

We define a gene to be diff. expressed if the adjusted p.value is $\leq `r fdr`$

```{r,include=TRUE,echo=FALSE,eval=TRUE,warning=FALSE,fig.width=9,fig.height=5}

diff_genes = tribble(
  ~ cell  , ~ results,
  "EBV",diff_genes_EBV,
  "NOKS",diff_genes_NOKS
) %>% 
  mutate(
    results = map(results,separate,gene_id,into = c("ENSEMBL","SYMBOL"),sep = "\\_"),
    diff_gene_list = map(results,filter,padj <= fdr) %>% 
      map(arrange,desc(stat)) %>% 
      map( ~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
        }),
    upreg_gene_list = map(results,filter,padj <= fdr) %>% 
      map(filter,log2FoldChange > 0) %>% 
      map(arrange, desc(stat)) %>% 
      map(~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
      }),
    downreg_gene_list = map(results,filter,padj <= fdr) %>% 
      map(filter,log2FoldChange < 0) %>% 
      map(arrange, desc(stat)) %>% 
      map(~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
      })
  )



diff_genes %>% 
  select(cell,results) %>% 
  unnest() %>% 
  ggplot(aes(-log10(padj)))+
  geom_histogram(bins = 100,fill = "white",colour = "black")+
  facet_grid( cell ~ . )+
  scale_x_continuous(limits = c(0,30))+
  geom_vline(xintercept = -log10(fdr),colour = "red",linetype =2)+
  theme_bw()


```





# Pathway analysis

This package appears to be contains several methods:

- An R made version of GSEA, we are using the GSEA model for a lists of genes with a signal-to-noise measure for each diff. expressed gene (which in this case is calculated by `r BiocStyle::Biocpkg("DESeq2")`.

- An over-representation test, which is a hyper-geometric test of the gene lists vs the genes in a certain pathway.

```{r path_analysis,message=FALSE,warning=FALSE}

pathway = diff_genes %>% 
  select(-results) %>% 
  gather(type,gene_list,-cell) %>% 
  mutate(
    type = str_replace(type,"_gene_list",""),
    type = if_else(type == "diff","all",type)
  ) %>% 
  mutate(
    gsea_analysis = map(gene_list,
                        GSEA,
                        TERM2GENE = pluck(hallmark,"symbols")),
    enricher_analysis = map(gene_list,names) %>% 
      map(enricher,TERM2GENE = pluck(hallmark,"symbols")),
    gsea_dotplots = map(gsea_analysis,dotplot),
    enricher_dotplots = map(enricher_analysis,dotplot))

```


## GSEA analysis

### Diff. expressed genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "all") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "all") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```

### Upregulated genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "upreg") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "upreg") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```

### Downregulated genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "downreg") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "downreg") %>% 
  pull(gsea_dotplots) %>% 
  pluck(1) 

```


## Over-representation test

### Diff. expressed genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "all") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1) 

```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "all") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1)

```

### Upregulated genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "upreg") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1) 

```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "upreg") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1) 

```

### Downregulated genes

- NOKS

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "NOKS") %>% 
  filter(type == "downreg") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1)
```

- EBV

```{r,include=TRUE,fig.width=9,fig.height=4}

pathway %>% 
  filter(cell == "EBV") %>% 
  filter(type == "downreg") %>% 
  pull(enricher_dotplots) %>% 
  pluck(1) 

```

## Comparison between EBV and NOKS

```{r over_rep}

comparison = pathway %>% 
  select(cell,type,gene_list) %>% 
  group_by(type) %>% 
  summarize(
    score_lists = list(gene_list)) %>% 
  ungroup() %>% 
  mutate(
    score_lists = map(score_lists,set_names,c("EBV","NOKS")),
    gene_lists = map(score_lists, ~ map(.,names))
  ) %>% 
  mutate(
    enricher_comp = map(gene_lists,compareCluster, fun = "enricher",
                            TERM2GENE = pluck(hallmark,"symbols")),
    enricher_dotplot = map(enricher_comp,dotplot)
  )


```

### All

```{r,fig.width=9,fig.height=5,include=TRUE}

comparison %>% 
  filter(type == "all") %>% 
  pull(enricher_dotplot) %>% 
  pluck(1)

```

### Upregulated

```{r,fig.width=9,fig.height=5,include=TRUE}

comparison %>% 
  filter(type == "upreg") %>% 
  pull(enricher_dotplot) %>% 
  pluck(1)

```


### Downregulated

```{r,fig.width=9,fig.height=5,include=TRUE}

comparison %>% 
  filter(type == "downreg") %>% 
  pull(enricher_dotplot) %>% 
  pluck(1)

```

## Comparison between upregulated and downregulated genes

```{r comp2}

comparison2 = pathway %>% 
  select(cell,type,gene_list) %>% 
  filter(type != "all") %>% 
  group_by(cell) %>% 
  summarize(
    score_lists = list(gene_list)) %>% 
  ungroup() %>% 
  mutate(
    score_lists = map(score_lists,set_names,c("Upregulated","Downregulated")),
    gene_lists = map(score_lists, ~ map(.,names))
  ) %>% 
  mutate(
    enricher_comp = map(gene_lists,compareCluster, fun = "enricher",
                            TERM2GENE = pluck(hallmark,"symbols")),
    enricher_dotplot = map(enricher_comp,dotplot)
  )

```

### EBV

```{r,fig.width=9,fig.height=5,include=TRUE}

comparison2 %>% 
  filter(cell == "EBV") %>% 
  pull(enricher_dotplot) %>% 
  pluck(1)

```

### NOKS

```{r,fig.width=9,fig.height=5,include=TRUE}

comparison2 %>% 
  filter(cell == "NOKS") %>% 
  pull(enricher_dotplot) %>% 
  pluck(1)

```



# Bibliography

G Yu, LG Wang, Y Han, QY He. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS: A Journal of Integrative Biology 2012, 16(5):284-287. doi:[10.1089/omi.2011.0118](http://dx.doi.org/10.1089/omi.2011.0118)

