---
title: "Enrichment analysis with clusterProfiler"
author: "Rene Welch"
date: February, 2018
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r start,include=FALSE,echo=FALSE,eval=TRUE}
## loads packages and data

library(org.Hs.eg.db)
library(tidyverse)
library(DESeq2)
library(clusterProfiler)
library(gridExtra)

theme_set(theme_bw())

fdr = 5e-2

knitr::opts_chunk$set(include = FALSE,echo = FALSE,eval = TRUE)

load(here::here("apps/Fig3_dashboad/fig3data.RData"))

```

# Introduction

We are going to perform a pathway analysis, I am using the data from Figure 3, i.e. genes that were differentiallly expressed in specific EBV or NOKS cell lines after applying the MC  treatment:

```{r data,include=TRUE}
ls()
```

# Preparing the data

## Gene Ontology list V5.1

We downloaded the **MolSigDB** curated gene sets from [the GSEA website](http://software.broadinstitute.org/gsea/downloads.jsp), and then we load the hallmark gene set from **MolSigDB**, using the 5.1 version:

```{r hallmark,cache=TRUE,include=TRUE,message=FALSE,warning=FALSE}

gmt_files = here::here("data/Pathway") %>% 
  list.files(full.names = TRUE,recursive = TRUE) %>% 
  str_subset("v5.1") %>% 
  str_subset(".gmt") %>% 
  str_subset("symbol")%>% 
  str_subset("c5") %>% 
  set_names(
    {
      basename(.) %>% 
        str_replace(".gmt","") %>% 
        str_split("\\.") %>% 
        map_chr( ~ .[2])
    }
  )

ontologies = gmt_files %>% 
  map(read.gmt)

gmt_files %>% 
  map_chr(basename)

```

There is one list of genes that contains all, and there are three separated subsets:

- __bp__: Biological processes
- __cc__: Cellular component
- __mf__: Molecular function

## Keratinocytes pathways

First, we search for ontologies with `KERA` in the name, and notice that those gene list are only present in the biological processes. So, we are going to focus on that gene list.


```{r onto_kera,include=TRUE,echo=TRUE}

onto_keratinocyte = ontologies %>% 
  map(filter,str_detect(ont,"KERA"))

onto_keratinocyte %>% 
  map_int(nrow)

onto_keratinocyte[["bp"]]

```

## Diff expressed genes

We define a gene to be diff. expressed if the adjusted p.value is $\leq `r fdr`$

```{r,include=TRUE,echo=FALSE,eval=TRUE,warning=FALSE,fig.width=9,fig.height=5}

diff_genes = tribble(
  ~ cell  , ~ results,
  "EBV",diff_genes_EBV,
  "NOKS",diff_genes_NOKS
) %>% 
  mutate(
    results = map(results,separate,gene_id,into = c("ENSEMBL","SYMBOL"),sep = "\\_"),
    diff_gene_list = map(results,filter,padj <= fdr) %>% 
      map(arrange,desc(stat)) %>% 
      map( ~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
        }),
    upreg_gene_list = map(results,filter,padj <= fdr) %>% 
      map(filter,log2FoldChange > 0) %>% 
      map(arrange, desc(stat)) %>% 
      map(~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
      }),
    downreg_gene_list = map(results,filter,padj <= fdr) %>% 
      map(filter,log2FoldChange < 0) %>% 
      map(arrange, desc(stat)) %>% 
      map(~ {
        mydata = .x
        pull(mydata,stat) %>% 
          set_names(pull(mydata,SYMBOL))
      })
  )



diff_genes %>% 
  select(cell,results) %>% 
  unnest() %>% 
  ggplot(aes(-log10(padj)))+
  geom_histogram(bins = 100,fill = "white",colour = "black")+
  facet_grid( cell ~ . )+
  scale_x_continuous(limits = c(0,30))+
  geom_vline(xintercept = -log10(fdr),colour = "red",linetype =2)+
  theme_bw()


```

# Pathway analysis on the Biological processes ontology

```{r path_analysis,message=FALSE,warning=FALSE,cache=TRUE}

pathway = diff_genes %>% 
  select(-results) %>% 
  gather(type,gene_list,-cell) %>% 
  mutate(
    type = str_replace(type,"_gene_list",""),
    type = if_else(type == "diff","all",type)
  ) %>% 
  mutate(
    gsea_0.05 = map(gene_list,
                        GSEA,
                        TERM2GENE = pluck(ontologies,"bp"),
                        pvalueCutoff = 0.05),
    gsea_0.1 = map(gene_list,
                         GSEA,
                         TERM2GENE = pluck(ontologies,"bp"),
                         pvalueCutoff = 0.1),
    enricher_0.05 = map(gene_list,names) %>% 
      map(enricher,TERM2GENE = pluck(ontologies,"bp"), 
          pvalueCutoff = 0.05),
    enricher_0.1 = map(gene_list,names) %>% 
      map(enricher,TERM2GENE = pluck(ontologies,"bp"),
          pvalueCutoff = 0.1))

extract_results = function(x){
  map(x,~.@result) %>% 
    map(as_tibble) %>% 
    map(mutate,
        log10pval = -log10(pvalue),
        log10padj = -log10(p.adjust))}

results = pathway %>% 
  select(-gene_list) %>% 
  mutate_if(negate(is.character),
            funs(extract_results(.)))
  
```

## Quick summary

For both methods `GSEA` and `enricher`, we performed the pathway analysis using the biological process subset of MolSigDB `C5: gene ontologies`. In total, for each cell and diff. expression category the following number of ontologies were deemed significant:

```{r summary,include=TRUE,echo=FALSE}

results %>% 
  mutate_if(is.list,funs( map_int(.,nrow) ))

```


```{r plot_funs}

barplot_NES = function(result, top_NES = 20, color_by = "p.adjust",
                       label_by = "setSize")
{
  
  pal = viridis::viridis(1000) %>% 
    rev()
  
  bb = result %>% 
    pull(NES) %>% 
    max()
  
  result %>% 
  top_n(n = top_NES,wt = NES) %>% 
    arrange(NES) %>% 
    mutate(ID = factor(ID,levels = ID),
           label_x = NES + 0.3) %>% 
    ggplot(aes_string("ID","NES",fill = color_by))+
    geom_bar(stat = "identity")+
    geom_text(aes_string("ID","label_x",label = label_by))+
    coord_flip()+
    scale_fill_gradientn(colors = pal)+
    theme(
      axis.title.y = element_blank()
    )+
    ylim(0, bb + .5)
    
}

```

```{r}
results %>% 
  pluck("gsea_0.05") %>% 
  pluck(4) %>% 
  barplot_NES(top_NES = 20, color_by = "log10padj",label_by = "setSize")

```



# Bibliography

G Yu, LG Wang, Y Han, QY He. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS: A Journal of Integrative Biology 2012, 16(5):284-287. doi:[10.1089/omi.2011.0118](http://dx.doi.org/10.1089/omi.2011.0118)

