---
title: "Enrichment analysis with clusterProfiler"
author: "Rene Welch"
date: February, 2018
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r start,include=FALSE,echo=FALSE,eval=TRUE}
## loads packages and data

library(org.Hs.eg.db)
library(tidyverse)
library(DESeq2)
library(clusterProfiler)
library(gridExtra)

knitr::opts_chunk$set(include = FALSE,echo = FALSE,eval = TRUE)

load(here::here("apps/Fig2_dashboad/fig2data.RData"))
fdr = 0.1


```

# Introduction

We are going to perform a pathway analysis, I am using the data from Figure 2, i.e. the genes that were differentially expressed between NOKS and EBV cell lines before any treatment was applied.

```{r data,include=TRUE}
ls()
```

Mark already analyzed the data using the TPM matrix and GSEA, so I am going to use `r BiocStyle::Biocpkg("clusterProfiler")` as an alternative method using the data already processed by `r BiocStyle::Biocpkg("DESeq2")`. 

As far as I can tell, the advantages of this package is that:

1. It can perform pathway analysis using the Hallmark gene dataset from **MolSigDB**

2. It contains several visualization methods, including some based on GSEA.

3. This may be minor, but for ChIP-seq annotation we have used `r BiocStyle::Biocpkg("ChIPseeker")` which was created by the same author.

# Pre-analysis

We downloaded the **MolSigDB** curated gene sets from [the GSEA website](http://software.broadinstitute.org/gsea/downloads.jsp), and then we load the hallmark gene set from **MolSigDB**.

```{r hallmark,cache=TRUE,include=TRUE,message=FALSE}

gmt_files = here::here("data/Pathway") %>% 
  list.files(full.names = TRUE) %>% 
  set_names(
    {
      basename(.) %>% 
        str_replace(".gmt","") %>% 
        str_split("\\.") %>% 
        map_chr( ~ .[length(.)])
    }
  )

gmt_files = gmt_files %>% 
  pluck("entrez") 

hallmark = gmt_files %>% 
  read.gmt() 

gmt_files

```

Lets recall from previous analysis, that for this dataset there are few differentially expressed genes, therefore we are using $\mbox{FDR} = `r fdr`. We are going to extract three gene sets of the list that we processed with `r BiocStyle::Biocpkg("DESeq2")`:

```{r genes,include=TRUE,echo=TRUE,warning=FALSE,message=FALSE}

diff_genes = diff_genes %>% 
  separate(gene_id,into = c("ENSEMBL","SYMBOL"), sep = "\\_") %>% 
  arrange(desc(stat))

ensembl = diff_genes %>% 
  pull(ENSEMBL) %>% 
  bitr(fromType = "ENSEMBL",toType = c("SYMBOL","ENTREZID"), OrgDb ="org.Hs.eg.db") %>% 
  as_tibble()

symbol = diff_genes %>% 
  pull(SYMBOL) %>% 
  bitr(fromType = "SYMBOL", toType = c("SYMBOL","ENTREZID"), OrgDb = "org.Hs.eg.db") %>% 
  as_tibble()

diff_genes = diff_genes %>% 
  right_join(ensembl,by = c("ENSEMBL","SYMBOL"))

```


```{r genes_entrez, include=TRUE,echo=TRUE,warning=FALSE,message=FALSE}
get_gene_entrez <- function(genes) {
  
  pull(genes,stat) %>% 
    set_names(
      pull(genes,ENTREZID)
    )
  
}

only_diff_genes = diff_genes  

gene_lists = tribble(
  ~ list_name, ~ genes_by_symbol,
  "all", 
    only_diff_genes %>% 
    get_gene_entrez(),
  "upregulated", 
    only_diff_genes %>%
    filter(log2FoldChange > 0) %>% 
    get_gene_entrez(),
  "downregulated",
    only_diff_genes %>% 
    filter(log2FoldChange < 0 ) %>% 
    get_gene_entrez()
)

gene_lists

```

## Sanity checks

It seems that only `r only_diff_genes %>% pull(ENTREZID) %>%  {. %in% hallmark$gene} %>% mean() %>% {round(.,4) *100}`% of our genes (labelled by ENTREZID) are in the hallmark dataset. Therefore, I need to check if the genome versions agree.

* First, do every gene SYMBOL has a ENSEMBL counterpart?

```{r symbol_ensembl_agree}






```



# Enrichment analysis


```{r}

##Note: I tried to use both the `enricher` and `GSEA` methods to perform pathway analysis using gene `SYMBOL`s. However, I wasn't able to use any of them. Probably due to the fact that only `r gene_lists$genes_by_symbol[[1]] %>% names %in% hallmark[[2]][,2] %>% mean %>% round(4)*100`% of the symbol names are in the hallmark gene set.

gene_lists$genes_by_symbol[[1]] %>% 
  enricher(TERM2GENE = hallmark)

## data(geneList, package="DOSE")





```


















# Bibliography

G Yu, LG Wang, Y Han, QY He. clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS: A Journal of Integrative Biology 2012, 16(5):284-287. doi:[10.1089/omi.2011.0118](http://dx.doi.org/10.1089/omi.2011.0118)

