---
title: "EBV and NOKS: Treated (MC) vs Untreated"
output:
  flexdashboard::flex_dashboard:
    theme: yeti
runtime: shiny
---

```{r global,include=TRUE,echo = FALSE,eval = TRUE}

load("fig3data.RData")

## block to check if the packages are installed
list.of.packages <- c("tidyverse","ggrepel","viridis","scales","shiny",
                      "rmarkdown","flexdashboard","d3heatmap",
                      "shinythemes","DT","RColorBrewer","hexbin",
                      "extrafontdb","pheatmap")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages))
  suppressMessages(install.packages(new.packages, repos = "http://cran.rstudio.com"))

library(tidyverse)
library(ggrepel)
library(viridis)
library(scales)
library(shiny)
library(d3heatmap)

my_alpha = .5
pal = viridis(1000)
my_style = "color: #FFFFFF; background-color: #FF0000"
my_style2 = "color: #FFFFFF; background-color: #357EC7"


theme_set(
  theme_bw()+
    theme(legend.position = "top",
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 16))
)

common_genes = diff_genes_EBV %>% 
  select(gene_id) %>% 
  inner_join(
    diff_genes_NOKS %>% select(gene_id), by = "gene_id"
  )

NOKS_vs_EBV_log2FC = rsem_data %>% 
  select(cell,treatment,rsem) %>%
  unnest() %>% 
  nest(-cell,-treatment) %>% 
  mutate(
    data = map(data,select,gene_id,TPM) %>% 
      map(group_by,gene_id) %>% 
      map(summarize,
          TPM = mean(TPM))
  ) %>% 
  unnest() %>% 
  nest(-cell)%>% 
  mutate(
    data = map(data,spread,treatment,TPM) %>% 
      map(mutate, log2FC = log2(1 + MC) - log2(1 + none)) %>% 
      map(select,gene_id,log2FC)
  ) %>% 
  unnest() %>% 
  spread(cell,log2FC) %>% 
  separate(gene_id , into = c("ensembl_id","gene_id"), sep = "\\_")

MA_TPM = rsem_data %>% 
  select(cell,treatment,rsem) %>% 
  unnest() %>% 
  group_by(cell,treatment,gene_id) %>% 
  summarize(
    TPM = mean(TPM)
  ) %>% 
  ungroup() %>% 
  nest(-cell,.key = cell_MA) %>% 
  mutate(
    cell_MA = map(cell_MA,spread,treatment,TPM)
  ) %>% 
  unnest() %>% 
  mutate_if(is.numeric,funs(log2(1 + .))) %>% 
  mutate(
    A = 0.5 * (MC + none),
    M = MC - none
  ) %>% 
  nest(-cell)

diff_genes = 
  tibble(
    cell = c("EBV","NOKS"),
    results = list(diff_genes_EBV,diff_genes_NOKS)
  ) %>% 
  inner_join(
    MA_TPM,by = "cell"
  ) %>% 
  mutate(
    results = map2(results,data, inner_join,by  = "gene_id"),
    data = NULL
  ) %>% 
  unnest() %>% 
  separate(gene_id,into = c("ensembl","gene_id"),sep = "\\_") %>% 
  nest(-cell,.key = results)

  EBV_genes = diff_genes %>% 
    filter(cell == "EBV") %>% 
    pluck("results") %>% 
    pluck(1)
  
  NOKS_genes = diff_genes %>% 
    filter(cell == "NOKS") %>% 
    pluck("results") %>% 
    pluck(1)

stri_subset = function(string,pattern)
{
  string[! str_detect(string,pattern)]
}

# tpmmat = rsem_data %>% 
#   select(file,rsem) %>% 
#   unnest() %>% 
#   select(file,gene_id,TPM) %>% 
#   spread(file,TPM) %>% 
#   as.data.frame() %>% 
#   tibble::remove_rownames() %>% 
#   tibble::column_to_rownames("gene_id") %>% 
#   as.matrix()

```
    
Inputs {.sidebar}
-------------------------------------

```{r sidebar}
# shiny inputs defined here

  numericInput("fdr","False Discovery Rate",1e-20,
               min = 1e-6,max = 1,step = .01)
  numericInput("ps","Point size",4)
  numericInput("ts","Axis ticks size",12)
  numericInput("Ts","Axis title size",16)
  numericInput("ls","Label size",6)
  hr()

  radioButtons("heatmap_type", "Heatmap type type",
            c("most DE", "selected"), inline = TRUE)

  numericInput("heatmap_most","Most DE genes",10,
               min = 5,max = 100,step = 1)
  
  hr()
  
  textInput("file",label = "Genes file", value = "fig3_genes.tsv")
  
  hr()
  textInput("loadfile",label = "Load genes",value = "fig3_genes.tsv")

  
  fillRow(flex = c(.4,.4),
          actionButton("load","Load",icon = icon("folder-open-o"),
                       style = my_style2),
          actionButton("save"," Save",icon = icon("floppy-o"),
                       style = my_style2),
          actionButton("clean"," Clean",icon = icon("window-close"),
                       style = my_style2))

  
    
  ## fix these
  observeEvent(input$save,{
    if(!is.null(ebv_genes$ensembl) | 
       !is.null(noks_genes$ensembl)){
      to_save = bind_rows(
        EBV_genes %>% 
          filter(ensembl %in% ebv_genes$ensembl) %>% 
          mutate(cell = "EBV"),
        NOKS_genes %>% 
          filter(ensembl %in% noks_genes$ensembl) %>% 
          mutate(cell = "NOKS")) %>% 
        select(cell,everything())

      write_tsv(to_save,input$file)
      
    }
  })
  
  observeEvent(input$load,{
    if(file.exists(input$loadfile)){
      diff_genes = read_tsv(input$loadfile)

      ebv = diff_genes %>% 
        filter(cell == "EBV") %>% 
        pluck("ensembl")
      
      noks = diff_genes %>% 
        filter(cell == "NOKS") %>% 
        pluck("ensembl")
      
      ebv_genes$ensembl = ebv  
      noks_genes$ensembl = noks  
      my_genes$ensembl = unique(c(ebv,noks))
    }
  })
  
  my_genes <- reactiveValues(ensembl = NULL)
  ebv_genes <- reactiveValues(ensembl = NULL)
  noks_genes <- reactiveValues(ensembl = NULL)
  
  gene_colors = c("black","darkblue","firebrick4")    

  observeEvent(input$clean,{
    ebv_genes$ensembl <- NULL
    noks_genes$ensembl <- NULL
  })
  
```

   
Column {.tabset}
-------------------------------------
   
### Alignment

```{r alignment}

alignment %>% 
    ggplot(
        aes(aligned,aligned_percent))+
    geom_point()+
    geom_text_repel(aes(label = replicate))+
    facet_grid(treat ~ cell)+
    theme(axis.text.x = element_text(angle = 25,
                                     hjust = 1))+
    scale_x_continuous(
        labels = comma)+
    scale_y_continuous(
        labels = percent)+
    xlab("Number of aligned reads")+
    ylab("Percentage of aligned reads")
```   

### p.value

```{r pval_histogram}

diff_genes %>% 
  unnest() %>% 
  ggplot(aes(pvalue))+
  geom_histogram(boundary = 0,
                 fill = "white",
                 color = "black",
                 bins = 51)+
  facet_grid( cell ~ .)+
  ylab("Number of genes")

```

### NOKS vs EBV 

```{r noks_vs_ebv_hexbin}

log2fc_ranges <- reactiveValues(x = NULL,y = NULL)

observeEvent(input$log2fc_dblclick, {
  brush <- input$log2fc_brush
  if (!is.null(brush)) {
    
    mm = min(brush$xmin,brush$ymin)
    MM = max(brush$xmax,brush$ymax)
    
    log2fc_ranges$x <- c(mm,MM)
    log2fc_ranges$y <- c(mm,MM)
  } else {
    log2fc_ranges$x <- NULL
    log2fc_ranges$y <- NULL
  }
})

observeEvent(input$log2fc_click,{
  ff <- nearPoints(NOKS_vs_EBV_log2FC,input$log2fc_click)
  ebv_genes$ensembl <- c(ebv_genes$ensembl, ff$ensembl_id)
  noks_genes$ensembl <- c(noks_genes$ensembl, ff$ensembl_id)
})

log2FC = reactive({

  both = intersect(ebv_genes$ensembl,
                   noks_genes$ensembl)
  only_ebv = setdiff(ebv_genes$ensembl,
                     noks_genes$ensembl)
  only_noks = setdiff(noks_genes$ensembl,
                      ebv_genes$ensembl)
  
  both_data = NOKS_vs_EBV_log2FC %>% 
    filter(ensembl_id %in% both)
  ebv_data = NOKS_vs_EBV_log2FC %>% 
    filter(ensembl_id %in% only_ebv)
  noks_data = NOKS_vs_EBV_log2FC %>% 
    filter(ensembl_id %in% only_noks)
  
  all_data = bind_rows(both_data %>% mutate(what = "both"),
                       ebv_data %>% mutate(what = "ebv"),
                       noks_data %>% mutate(what = "noks"))

  NOKS_vs_EBV_log2FC %>%
    ggplot(aes(NOKS,EBV))+
    stat_binhex(bins = 101) +
    scale_fill_gradientn(colours = pal,trans = "log10",
                         labels = trans_format('log10',math_format(10^.x)),
                         guide = guide_colourbar(title = NULL,
                           barheight = unit(0.92,"npc"),
                           barwidth = unit(0.01,"npc")))+
  theme(legend.position = "right",
        axis.text = element_text(size = input$ts),
        axis.title = element_text(size = input$Ts),
        legend.text = element_text(size = input$ts),
        legend.title = element_text(size = input$ts))+
  coord_cartesian(
    xlim = log2fc_ranges$x,
    ylim = log2fc_ranges$y)+
    geom_abline(slope = 1,intercept = 0,linetype =2)+
    geom_vline(xintercept = 0,linetype = 2,colour = "red")+
    geom_hline(yintercept = 0,linetype = 2,colour = "red")+
    geom_label_repel(data = all_data,aes(colour = what,label = gene_id),
                     box.padding = 0.35,
                     point.padding = 0.5,
                     segment.colour = "grey50",size = input$ls)+
    scale_color_manual(values = gene_colors)+
    guides(colour = FALSE)

})


output$log2fc = renderPlot(log2FC())
 
fillCol(flex = c(1,.1),
        plotOutput("log2fc",
                   click = "log2fc_click",
                   dblclick = "log2fc_dblclick",
                   brush = brushOpts(
                     id = "log2fc_brush",
                     resetOnNew = TRUE
                     )),
        fillRow(flex = c(.1,.3,.3,.3),
          actionButton("log2fc_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("log2fc_width","Log2FC width (in)",7),
          numericInput("log2fc_height","Log2FC height (in)",7),
          textInput("log2fc_file","Log2FC file","log2fc_file.pdf")))

observeEvent(input$log2fc_save,{
  
  ggsave(
    plot = log2FC(),
    filename = input$log2fc_file,
    width = unit(input$log2fc_width,"in"),
    height = unit(input$log2fc_height,"in")
  )})



```

### Volcano EBV 
    
```{r volcano_ebv}

volcano_ebv_ranges <- reactiveValues(x = NULL,y = NULL)

observeEvent(input$volcano_ebv_dblclick, {
  brush <- input$volcano_ebv_brush
  if (!is.null(brush)) {
    volcano_ebv_ranges$x <- c(brush$xmin, brush$xmax)
    volcano_ebv_ranges$y <- c(brush$ymin, brush$ymax)
    volcano_ebv_ranges$y[1] <- max(volcano_ebv_ranges$y[1],0)
  } else {
    volcano_ebv_ranges$x <- NULL
    volcano_ebv_ranges$y <- NULL
  }
})

observeEvent(input$volcano_ebv_click,{
  ff <- nearPoints(EBV_genes,input$volcano_ebv_click)
  ebv_genes$ensembl <- c(ebv_genes$ensembl, ff$ensembl)
})

volcano_EBV = reactive({

  both = intersect(ebv_genes$ensembl,
                   noks_genes$ensembl)
  only_ebv = setdiff(ebv_genes$ensembl,
                     noks_genes$ensembl)
  only_noks = setdiff(noks_genes$ensembl,
                      ebv_genes$ensembl)
  
  both_data = EBV_genes %>% 
    filter(ensembl %in% both)
  
  ebv_data = EBV_genes %>%
    filter(ensembl %in% only_ebv)
  
  noks_data = EBV_genes %>% 
    filter(ensembl %in% only_noks)

  all_data = bind_rows(both_data %>% mutate(what = "both"),
                       ebv_data %>% mutate(what = "ebv"),
                       noks_data %>% mutate(what = "noks"))

  
  EBV_genes %>%
    ggplot(aes(log2FoldChange,log10pval,
               colour = if_else(padj <= input$fdr,"yes","no")))+
    geom_point(size = input$ps,alpha = .75)+
    geom_label_repel(data = all_data,aes(fill = what,label = gene_id),
                     colour = "white",box.padding = 0.35,
                     point.padding = 0.5,
                     segment.colour = "grey50",size = input$ls)+
    scale_fill_manual(values = gene_colors)+
    guides(fill = FALSE)+
    scale_color_brewer(palette = "Set1",name = "Diff. Expressed")+
    coord_cartesian(
      xlim = volcano_ebv_ranges$x ,
      ylim = volcano_ebv_ranges$y
    )+
    theme(
        axis.text = element_text(size = input$ts),
        axis.title = element_text(size = input$Ts),
        legend.text = element_text(size = input$ts),
        legend.title = element_text(size = input$ts))+
    ylab("-log10(pvalue)")+
    geom_vline(xintercept = 0,colour = "black",linetype = 2)

})

output$volcano_ebv = renderPlot(volcano_EBV())
 
fillCol(flex = c(1,.1),
        plotOutput("volcano_ebv",
                   click = "volcano_ebv_click",
                   dblclick = "volcano_ebv_dblclick",
                   brush = brushOpts(
                     id = "volcano_ebv_brush",
                     resetOnNew = TRUE
                     )),
        fillRow(flex = c(.1,.3,.3,.3),
          actionButton("volcano_ebv_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("volcano_ebv_width","Volcano EBV width (in)",7),
          numericInput("volcano_ebv_height","Volcano EBV height (in)",7),
          textInput("volcano_ebv_file","Volcano EBV file","volcano_ebv_file.pdf")))

observeEvent(input$volcano_ebv_save,{
  
  ggsave(
    plot = volcano_EBV(),
    filename = input$volcano_ebv_file,
    width = unit(input$volcano_ebv_width,"in"),
    height = unit(input$volcano_ebv_height,"in")
  )})




```

### Volcano NOKS

```{r volcano_noks}

volcano_noks_ranges <- reactiveValues(x = NULL,y = NULL)

observeEvent(input$volcano_noks_dblclick, {
  brush <- input$volcano_noks_brush
  if (!is.null(brush)) {
    volcano_noks_ranges$x <- c(brush$xmin, brush$xmax)
    volcano_noks_ranges$y <- c(brush$ymin, brush$ymax)
    volcano_noks_ranges$y[1] <- max(volcano_noks_ranges$y[1],0)
  } else {
    volcano_noks_ranges$x <- NULL
    volcano_noks_ranges$y <- NULL
  }
})

observeEvent(input$volcano_noks_click,{
  ff <- nearPoints(NOKS_genes,input$volcano_noks_click)
  noks_genes$ensembl <- c(noks_genes$ensembl, ff$ensembl)
})

volcano_NOKS = reactive({

  both = intersect(ebv_genes$ensembl,
                   noks_genes$ensembl)
  only_ebv = setdiff(ebv_genes$ensembl,
                     noks_genes$ensembl)
  only_noks = setdiff(noks_genes$ensembl,
                      ebv_genes$ensembl)
  
  both_data = NOKS_genes %>% 
    filter(ensembl %in% both)
  ebv_data = NOKS_genes %>%
    filter(ensembl %in% only_ebv)
  noks_data = NOKS_genes %>% 
    filter(ensembl %in% only_noks)

  all_data = bind_rows(both_data %>% mutate(what = "both"),
                       ebv_data %>% mutate(what = "ebv"),
                       noks_data %>% mutate(what = "noks"))

  
  NOKS_genes %>%
    ggplot(aes(log2FoldChange,log10pval,
               colour = if_else(padj <= input$fdr,"yes","no")))+
    geom_point(size = input$ps,alpha = .75)+
    geom_label_repel(data = all_data,aes(fill = what,label = gene_id),
                     colour = "white",box.padding = 0.35,
                     point.padding = 0.5,
                     segment.colour = "grey50",size = input$ls)+
    scale_fill_manual(values = gene_colors)+
    guides(fill = FALSE)+
    scale_color_brewer(palette = "Set1",name = "Diff. Expressed")+
    coord_cartesian(
      xlim = volcano_noks_ranges$x ,
      ylim = volcano_noks_ranges$y
    )+
    ylab("-log10(pvalue)")+
    theme(
        axis.text = element_text(size = input$ts),
        axis.title = element_text(size = input$Ts),
        legend.text = element_text(size = input$ts),
        legend.title = element_text(size = input$ts))+
    geom_vline(xintercept = 0,colour = "black",linetype = 2)
  
})

output$volcano_noks = renderPlot(volcano_NOKS())
 
fillCol(flex = c(1,.1),
        plotOutput("volcano_noks",
                   click = "volcano_noks_click",
                   dblclick = "volcano_noks_dblclick",
                   brush = brushOpts(
                     id = "volcano_noks_brush",
                     resetOnNew = TRUE
                     )),
        fillRow(flex = c(.1,.3,.3,.3),
          actionButton("volcano_noks_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("volcano_noks_width","Volcano NOKS width (in)",7),
          numericInput("volcano_noks_height","Volcano NOKS height (in)",7),
          textInput("volcano_noks_file","Volcano NOKS file","volcano_noks_file.pdf")))

observeEvent(input$volcano_noks_save,{
  
  ggsave(
    plot = volcano_NOKS(),
    filename = input$volcano_noks_file,
    width = unit(input$volcano_noks_width,"in"),
    height = unit(input$volcano_noks_height,"in")
  )})
        


```

### MA EBV

```{r ma_ebv}

ma_ebv_ranges <- reactiveValues(x = NULL,y = NULL)

observeEvent(input$ma_ebv_dblclick, {
  brush <- input$ma_ebv_brush
  if (!is.null(brush)) {
    ma_ebv_ranges$x <- c(brush$xmin, brush$xmax)
    ma_ebv_ranges$y <- c(brush$ymin, brush$ymax)
    ma_ebv_ranges$x[1] <- max(ma_ebv_ranges$x[1],0)
  } else {
    ma_ebv_ranges$x <- NULL
    ma_ebv_ranges$y <- NULL
  }
})

observeEvent(input$ma_ebv_click,{
  ff <- nearPoints(EBV_genes,input$ma_ebv_click)
  ebv_genes$ensembl <- c(ebv_genes$ensembl, ff$ensembl)
})

MA_EBV = reactive({

  both = intersect(ebv_genes$ensembl,
                   noks_genes$ensembl)
  only_ebv = setdiff(ebv_genes$ensembl,
                     noks_genes$ensembl)
  only_noks = setdiff(noks_genes$ensembl,
                      ebv_genes$ensembl)
  
  both_data = EBV_genes %>% 
    filter(ensembl %in% both)
  
  ebv_data = EBV_genes %>%
    filter(ensembl %in% only_ebv)
  
  noks_data = EBV_genes %>% 
    filter(ensembl %in% only_noks)

  all_data = bind_rows(both_data %>% mutate(what = "both"),
                       ebv_data %>% mutate(what = "ebv"),
                       noks_data %>% mutate(what = "noks"))
  
  
  EBV_genes %>%
    ggplot(aes(A,M,
               colour = if_else(padj <= input$fdr,"yes","no")))+
    geom_point(size = input$ps,alpha = .75)+
    geom_label_repel(data = all_data,aes(fill = what,label = gene_id),
                     colour = "white",box.padding = 0.35,
                     point.padding = 0.5,
                     segment.colour = "grey50",size = input$ls)+
    scale_fill_manual(values = gene_colors)+
    guides(fill = FALSE)+
    scale_color_brewer(palette = "Set1",name = "Diff. Expressed")+
    coord_cartesian(
      xlim = ma_ebv_ranges$x ,
      ylim = ma_ebv_ranges$y
    )+
    theme(
        axis.text = element_text(size = input$ts),
        axis.title = element_text(size = input$Ts),
        legend.text = element_text(size = input$ts),
        legend.title = element_text(size = input$ts))+
    geom_hline(yintercept = 0,colour = "black",linetype = 2)
  
})

output$ma_ebv = renderPlot(MA_EBV())
 
fillCol(flex = c(1,.1),
        plotOutput("ma_ebv",
                   click = "ma_ebv_click",
                   dblclick = "ma_ebv_dblclick",
                   brush = brushOpts(
                     id = "ma_ebv_brush",
                     resetOnNew = TRUE
                     )),
        fillRow(flex = c(.1,.3,.3,.3),
          actionButton("ma_ebv_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("ma_ebv_width","MA EBV width (in)",7),
          numericInput("ma_ebv_height","MA EBV height (in)",7),
          textInput("ma_ebv_file","MA EBV file","ma_ebv_file.pdf")))

observeEvent(input$ma_ebv_save,{
  
  ggsave(
    plot = MA_EBV(),
    filename = input$ma_ebv_file,
    width = unit(input$ma_ebv_width,"in"),
    height = unit(input$ma_ebv_height,"in")
  )})


```

### MA NOKS

```{r ma_noks}

ma_noks_ranges <- reactiveValues(x = NULL,y = NULL)

observeEvent(input$ma_noks_dblclick, {
  brush <- input$ma_noks_brush
  if (!is.null(brush)) {
    ma_noks_ranges$x <- c(brush$xmin, brush$xmax)
    ma_noks_ranges$y <- c(brush$ymin, brush$ymax)
    ma_noks_ranges$x[1] <- max(ma_noks_ranges$x[1],0)
  } else {
    ma_noks_ranges$x <- NULL
    ma_noks_ranges$y <- NULL
  }
})

observeEvent(input$ma_noks_click,{
  ff <- nearPoints(NOKS_genes,input$ma_noks_click)
  noks_genes$ensembl <- c(noks_genes$ensembl, ff$ensembl)
})

MA_NOKS = reactive({

  both = intersect(ebv_genes$ensembl,
                   noks_genes$ensembl)
  only_ebv = setdiff(ebv_genes$ensembl,
                     noks_genes$ensembl)
  only_noks = setdiff(noks_genes$ensembl,
                      ebv_genes$ensembl)
  
  both_data = NOKS_genes %>% 
    filter(ensembl %in% both)
  ebv_data = NOKS_genes %>%
    filter(ensembl %in% only_ebv)
  noks_data = NOKS_genes %>% 
    filter(ensembl %in% only_noks)

  all_data = bind_rows(both_data %>% mutate(what = "both"),
                       ebv_data %>% mutate(what = "ebv"),
                       noks_data %>% mutate(what = "noks"))

  
  NOKS_genes %>%
    ggplot(aes(A,M,
               colour = if_else(padj <= input$fdr,"yes","no")))+
    geom_point(size = input$ps,alpha = .75)+
    geom_label_repel(data = all_data,aes(fill = what,label = gene_id),
                     colour = "white",box.padding = 0.35,
                     point.padding = 0.5,
                     segment.colour = "grey50",size = input$ls)+
    scale_fill_manual(values = gene_colors)+
    guides(fill = FALSE)+
    scale_color_brewer(palette = "Set1",name = "Diff. Expressed")+
    coord_cartesian(
      xlim = ma_noks_ranges$x ,
      ylim = ma_noks_ranges$y
    )+
    theme(
        axis.text = element_text(size = input$ts),
        axis.title = element_text(size = input$Ts),
        legend.text = element_text(size = input$ts),
        legend.title = element_text(size = input$ts))+
    geom_hline(yintercept = 0,colour = "black",linetype = 2)
  
  
})

output$ma_noks = renderPlot(MA_NOKS())
 
fillCol(flex = c(1,.1),
        plotOutput("ma_noks",
                   click = "ma_noks_click",
                   dblclick = "ma_noks_dblclick",
                   brush = brushOpts(
                     id = "ma_noks_brush",
                     resetOnNew = TRUE
                     )),
        fillRow(flex = c(.1,.3,.3,.3),
          actionButton("ma_noks_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("ma_noks_width","MA NOKS width (in)",7),
          numericInput("ma_noks_height","MA NOKS height (in)",7),
          textInput("ma_noks_file","MA NOKS file","ma_noks_file.pdf")))

observeEvent(input$ma_noks_save,{
  
  ggsave(
    plot = MA_NOKS(),
    filename = input$ma_noks_file,
    width = unit(input$ma_noks_width,"in"),
    height = unit(input$ma_noks_height,"in")
  )})        


```

### EBV heatmap

```{r ebv_heatmap}

ebv_mat = reactive({

  my_mat = rlogmat

  if(input$heatmap_type == "most DE"){
    genes = EBV_genes %>%
      arrange(desc(log10pval)) %>%
      head(input$heatmap_most )

      idx = genes %>%
        mutate(
          idx = paste(ensembl,gene_id,sep = "_")
        ) %>%
        pluck("idx")
      
      if(length(idx) <= 1){
        my_mat = my_mat[1:2,]
      }else{
        my_mat = my_mat[rownames(my_mat) %in% idx, ]
      }

  }else{
    if(is.null(ebv_genes$ensembl)){
      my_mat = my_mat[sample(nrow(my_mat),3),]
    }else{
      genes = EBV_genes %>%
        filter(ensembl %in% ebv_genes$ensembl)

      idx = genes %>%
        mutate(
          idx = paste(ensembl,gene_id,sep = "_")
        ) %>%
        pluck("idx")
      my_mat = my_mat[rownames(my_mat) %in% idx, ]
    }

  }
  rownames(my_mat) = rownames(my_mat) %>%
    str_split("_") %>%
    map_chr( ~ .[2])

  colnames(my_mat) = colnames(my_mat) %>%
    str_replace("RNAseq-","")
  
  my_mat

})

fillCol(flex = c(1,.1),
        renderD3heatmap({
          
          my_mat = ebv_mat()
          pal = viridis(100)
          
          d3heatmap(my_mat,
                    colors = pal,
                    xaxis_font_size = 12,
                    yaxis_font_size = 12)}),
        fillRow(flex = c(.1,.3,.3,.3),
                actionButton("ebv_heat_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("ebv_heat_width","EBV Heatmap width (in)",7),
          numericInput("ebv_heat_height","EBV Heatmap height (in)",7),
          textInput("ebv_heat_file","EBV Heatmap file","ebvh_file.pdf")))

observeEvent(input$ebv_heat_save,{

  pal = viridis(100)
  
  pheatmap::pheatmap(
    mat = ebv_mat(),
    scale = "row",
    color = pal,
    fontsize = input$ts,
    width = input$ebv_heat_width,
    height = input$ebv_heat_height,
    filename = input$ebv_heat_file
  )

})


```

### NOKS heatmap

```{r noks_heatmap}

noks_mat = reactive({
  
  my_mat = rlogmat

  if(input$heatmap_type == "most DE"){
    genes = NOKS_genes %>%
      arrange(desc(log10pval)) %>%
      head(input$heatmap_most )

      idx = genes %>%
        mutate(
          idx = paste(ensembl,gene_id,sep = "_")
        ) %>%
        pluck("idx")

      if(length(idx) <= 1){
        my_mat = my_mat[1:2,]
      }else{
        my_mat = my_mat[rownames(my_mat) %in% idx, ]
      }

  }else{
    if(is.null(noks_genes$ensembl)){
      my_mat = my_mat[sample(nrow(my_mat),3),]
    }else{
      genes = NOKS_genes %>%
        filter(ensembl %in% noks_genes$ensembl)

      idx = genes %>%
        mutate(
          idx = paste(ensembl,gene_id,sep = "_")
        ) %>%
        pluck("idx")
      my_mat = my_mat[rownames(my_mat) %in% idx, ]
    }

  }
  rownames(my_mat) = rownames(my_mat) %>%
    str_split("_") %>%
    map_chr( ~ .[2])

  colnames(my_mat) = colnames(my_mat) %>%
    str_replace("RNAseq-","")

  my_mat
})




fillCol(flex = c(1,.1),
        renderD3heatmap({
          
          my_mat = noks_mat()
          pal = viridis(100)
          
          d3heatmap(my_mat,
                    colors = pal,
                    xaxis_font_size = 12,
                    yaxis_font_size = 12)}),
        fillRow(flex = c(.1,.3,.3,.3),
                actionButton("noks_heat_save","",icon = icon("camera"),
                       style = my_style),
          numericInput("noks_heat_width","NOKS Heatmap width (in)",7),
          numericInput("noks_heat_height","NOKS Heatmap height (in)",7),
          textInput("noks_heat_file","NOKS Heatmap file","noksh_file.pdf")))

observeEvent(input$noks_heat_save,{

  pal = viridis(100)
  
  pheatmap::pheatmap(
    mat = noks_mat(),
    scale = "row",
    color = pal,
    fontsize = input$ts,
    width = input$noks_heat_width,
    height = input$noks_heat_height,
    filename = input$noks_heat_file
  )

})




```


### EBV genes 

```{r ebv_genes}  

div(
  renderDataTable(
    EBV_genes %>%
      filter(ensembl %in% ebv_genes$ensembl) %>%
      select(gene_id,ensembl,padj,log10pval,MC,none,stat),
    options = list(
      pageLength = 10,
      lengthMenu = c(10,20,50,100)
    ) ) ,style = "font-size:80%")

```

### NOKS genes

```{r noks_genes}  

div(
  renderDataTable(
    NOKS_genes %>%
      filter(ensembl %in% noks_genes$ensembl) %>%
      select(gene_id,ensembl,padj,log10pval,MC,none,stat),
    options = list(
      pageLength = 10,
      lengthMenu = c(10,20,50,100)
    ) ) ,style = "font-size:80%")

```

### Instructions

**Alignment** - This plot is fixed, I think for the supplement

**p.values** - Also fixed, there is no deviation of the assumptions. Both look like they are supposed: a peak at zero and then approx. uniform. A quick note, is that the peak at zero is really tall, which means that there are a lot of diff. expressed genes. We may want to use smaller values for setting the FDR (by default 1e-20).

**volcano** and **MA plots** - In these plots, can click the point and build a table of DE genes. To change the scale of the figure: brush and double click. If you double click without brush that will restart to the initial plot. Just click on a gene to label it, the results are stored in both **heatmaps** and **Gene table**. Removed the MA-DESeq plot because it didn't seem very useful.

**heatmaps** The TPM heatmaps from fig2 were hard to distinguish, so I tought it would be better to remove them, and only keep the model ones (is the same model for both, the differences are going to be the p.value used to rank or the genes selected)

**tables** They are separated for display, but the `save genes` button will save everything in the genes file

**Sidebar**

* False discovery rate: The smaller, the test results are more conservative

* Point size: Just to make the plots to look just right

* Clean gene lists: To start over again (changed to have one per cell)

* Save genes and file: When some genes are selected from the volcano or MA plots,this button will save the list in the file



